Regional Base-Flow Index in Arid Landscapes Using Machine Learning and Instrumented Records

### 2025-03-26 elsevier_MANUSCRIPT.qmd
updated

| **Classification Group** | **n** | **n_pos** | **n_neg** | **perc_pos** | **perc_neg** |
|-------------|------------|------------|------------|------------|------------|
| Precipitation - Monsoon Dominated | 87 | 8 | 21 | 0.092 | 0.241 |
| Precipitation - Snowmelt Dominated | 118 | 9 | 12 | 0.076 | 0.102 |
| Physiographic Region - Basin and Range | 156 | 14 | 26 | 0.090 | 0.167 |
| Physiographic Region - Colorado Plateau | 49 | 3 | 7 | 0.061 | 0.143 |
| Climate - Warm-Wet | 31 | 2 | 6 | 0.065 | 0.194 |
| Climate - Warm-Dry | 55 | 6 | 11 | 0.109 | 0.200 |
| Climate - Cool-Wet | 74 | 4 | 9 | 0.054 | 0.122 |
| Climate - Cool-Dry | 45 | 5 | 7 | 0.111 | 0.156 |
| Slope - High | 102 | 10 | 12 | 0.098 | 0.118 |
| Slope - Low | 103 | 7 | 21 | 0.068 | 0.204 |

@@

| **Classification Group** | **n** | **R2** | **MSE** | **RMSE** | **MAE** | **NSE** | **pbias** |
|---------|---------|---------|---------|---------|---------|---------|---------|
| Climate - Monsoon Dominated | 3039 | 0.633 | 0.016 | 0.126 | 0.074 | 0.619 | -13.7 |
| Climate - Snowmelt Dominated | 4685 | 0.733 | 0.015 | 0.121 | 0.087 | 0.725 | -3.5 |
| PhysRegion - Basin&Range | 6147 | 0.733 | 0.016 | 0.127 | 0.084 | 0.724 | -6.3 |
| PhysRegion - CO Plateau | 1577 | 0.846 | 0.011 | 0.104 | 0.073 | 0.843 | -3.8 |
| Climate - Warm-Wet | 1506 | 0.693 | 0.014 | 0.117 | 0.077 | 0.685 | -8.2 |
| Climate - Warm-Dry | 2351 | 0.693 | 0.022 | 0.147 | 0.092 | 0.675 | -11.9 |
| Climate - Cool-Wet | 2350 | 0.738 | 0.011 | 0.106 | 0.078 | 0.736 | -1.7 |
| Climate - Cool-Dry | 1517 | 0.831 | 0.012 | 0.111 | 0.078 | 0.827 | -4.3 |
| Slope - High | 3795 | 0.776 | 0.012 | 0.111 | 0.079 | 0.771 | -3.3 |
| Slope - Low | 3929 | 0.724 | 0.018 | 0.133 | 0.085 | 0.713 | -9.1 |

@@

Previous studies have examined base-flow regionalization and synthesis across various spatial scales, some overlapping with our study area. @beck2013 explored global patterns of BFI using streamflow observations extrapolated through an artificial neural network (ANN). The U.S. streamgages included in that analysis were drawn from the interim Model Parameter Estimation Experiment (MOPEX) dataset of 1862 USGS gauges rather than the finalized dataset of 431 gauges meeting precipitation gauge requirements [@beck2013; @mopex_2006]. Only 10 of these interim gauges overlap geographically with our study area. Similarly, @ayers2022 assessed climatic influences on monthly baseflow across the continental United States (CONUS), utilizing a subset of GAGES-II streamflow gauges. Although exact gauge locations from that study are unavailable, published figures suggest minimal geographic overlap with our research domain. @santhi2008 employed interpolation methods for regional base-flow estimation across CONUS, but specific gauge locations or data are not publicly available, limiting direct regional comparisons.

The utility of existing streamgage networks for base-flow analyses in Arizona is limited by their design and monitoring objectives. While the number of streamgages has increased to meet regulatory imperatives, such as those driven by the Clean Water Act, many gauges prioritize peak flow monitoring rather than low-flow conditions critical for base-flow studies [@flood-control2020]. Newer non-USGS gauges, such as those installed by flood control districts (e.g., the ALERT system), are generally tailored for flood detection rather than monitoring sustained base flow. Recent USGS installations focus on addressing in-stream flow rights and future monitoring improvements, though low-flow capabilities remain limited.

This study addresses these monitoring gaps through a detailed analysis of observed base flow and BFI at instrumented streamgages, highlighting groundwater-surface water interactions within Arizona’s arid and semi-arid landscapes. Observed spatial patterns in BFI clearly align with known hydrogeological and climatic variations. High BFI values coincide with areas characterized by spring-fed streams and pronounced physiographic transitions, such as the Grand Canyon and Mogollon Rim. Conversely, lower observed BFI values correspond to regions with limited groundwater contributions, notably the Defiance Plateau and southern Arizona, emphasizing vulnerability in these arid environments.

@@

Given the limited coverage provided by existing gauges, a common issue in arid and semi-arid regions globally [@krabbenhoft-2022], we utilized a machine learning approach to estimate BFI in ungauged catchments. The regional model exhibited strong predictive capability (overall R²=0.764), though systematic underprediction (negaitve pbias) was noted, particularly in monsoon-dominated and warm-dry climates. These biases suggest areas for improvement in model performance, potentially through the incorporation of additional predictors or more refined regional hydrologic representations. Although creating separate regional models (by physiographic region) was explored to enhance predictive accuracy, the statewide model consistently outperformed these regionalized approaches, emphasizing the advantage of a comprehensive, statewide dataset.

Elevation emerged as the most significant predictor, reinforcing its established role as a proxy for climate conditions influencing groundwater recharge. This aligns with findings from previous studies conducted in diverse landscapes, including New Zealand [@singh2018] and across the conterminous U.S. [@santhi2008; @beck2013]. The mixed influence observed for Soil Type A indicates complexity in infiltration dynamics, suggesting that future research could benefit from explicitly incorporating temporal variables such as snowpack duration, soil moisture, and detailed land-cover characteristics.

@@

This research highlights critical limitations in Arizona’s current streamgage network for monitoring base-flow processes, suggesting a need for targeted investments in instrumentation capable of capturing low-flow dynamics. Additionally, the demonstrated success of our modeling framework underscores its broader applicability to similarly data-scarce dryland regions globally, providing a valuable tool for water resource management and climate adaptation strategies. By improving our understanding of groundwater-surface water interactions under changing climatic conditions, this study contributes substantially to addressing hydrological uncertainties in arid and semi-arid environments.

>>>>

| **Classification Group** | **n** | **n_pos** | **n_neg** | **perc_pos** | **perc_neg** |
|---------------------|----------|----------|----------|----------|----------|
| Precipitation - Monsoon Dominated | 87 | 8 | 21 | 0.092 | 0.241 |
| Precipitation - Snowmelt Dominated | 118 | 9 | 12 | 0.076 | 0.102 |
| Physiographic Region - Basin and Range | 156 | 14 | 26 | 0.090 | 0.167 |
| Physiographic Region - Colorado Plateau | 49 | 3 | 7 | 0.061 | 0.143 |
| Climate - Warm-Wet | 31 | 2 | 6 | 0.065 | 0.194 |
| Climate - Warm-Dry | 55 | 6 | 11 | 0.109 | 0.200 |
| Climate - Cool-Wet | 74 | 4 | 9 | 0.054 | 0.122 |
| Climate - Cool-Dry | 45 | 5 | 7 | 0.111 | 0.156 |
| Slope - High | 102 | 10 | 12 | 0.098 | 0.118 |
| Slope - Low | 103 | 7 | 21 | 0.068 | 0.204 |

@@

| **Classification Group** | **n** | **R2** | **MSE** | **RMSE** | **MAE** | **NSE** | **pbias** |
|-------------|---------|---------|---------|---------|---------|---------|---------|
| Climate - Monsoon Dominated | 3039 | 0.633 | 0.016 | 0.126 | 0.074 | 0.619 | -13.7 |
| Climate - Snowmelt Dominated | 4685 | 0.733 | 0.015 | 0.121 | 0.087 | 0.725 | -3.5 |
| PhysRegion - Basin&Range | 6147 | 0.733 | 0.016 | 0.127 | 0.084 | 0.724 | -6.3 |
| PhysRegion - CO Plateau | 1577 | 0.846 | 0.011 | 0.104 | 0.073 | 0.843 | -3.8 |
| Climate - Warm-Wet | 1506 | 0.693 | 0.014 | 0.117 | 0.077 | 0.685 | -8.2 |
| Climate - Warm-Dry | 2351 | 0.693 | 0.022 | 0.147 | 0.092 | 0.675 | -11.9 |
| Climate - Cool-Wet | 2350 | 0.738 | 0.011 | 0.106 | 0.078 | 0.736 | -1.7 |
| Climate - Cool-Dry | 1517 | 0.831 | 0.012 | 0.111 | 0.078 | 0.827 | -4.3 |
| Slope - High | 3795 | 0.776 | 0.012 | 0.111 | 0.079 | 0.771 | -3.3 |
| Slope - Low | 3929 | 0.724 | 0.018 | 0.133 | 0.085 | 0.713 | -9.1 |

@@

Previous studies have examined base-flow regionalization and synthesis across various spatial scales, some overlapping with our study area. @beck2013 explored global patterns of BFI using streamflow observations extrapolated through an artificial neural network (ANN). The U.S. streamgages included in that analysis were drawn from the interim Model Parameter Estimation Experiment (MOPEX) dataset of 1862 USGS gauges [@beck2013; @mopex_2006]. Only 10 of these interim gauges overlap geographically with our study area. Similarly, @ayers2022 assessed climatic influences on monthly baseflow across the continental United States (CONUS), utilizing a subset of GAGES-II streamflow gauges. Although exact gauge locations from that study are unavailable, published figures suggest minimal geographic overlap with our research domain. @santhi2008 employed interpolation methods for regional base-flow estimation across CONUS, but specific gauge locations or data are not available, limiting direct regional comparisons.

The existing streamgage networks Arizona are limited for base-flow analyses by their design and monitoring objectives. While the number of streamgages has increased to meet regulatory imperatives, such as those driven by the Clean Water Act, many gauges prioritize peak flow monitoring rather than low-flow conditions critical for base-flow studies [@flood-control2020]. Newer non-USGS gauges, such as those installed by flood control districts (e.g., the ALERT system), are generally tailored for flood detection rather than monitoring sustained base flow. Recent USGS installations focus on addressing in-stream flow rights and future monitoring improvements, though low-flow capabilities remain limited.

This study addresses these monitoring gaps through an analysis of observed base flow and BFI at instrumented streamgages, highlighting groundwater-surface water interactions within Arizona’s arid and semi-arid landscapes. Observed spatial patterns in BFI clearly align with known hydrogeological and climatic variations. High BFI values coincide with areas characterized by spring-fed streams and pronounced physiographic transitions, such as the Grand Canyon and Mogollon Rim. Conversely, lower observed BFI values correspond to regions with limited groundwater contributions, notably the Defiance Plateau and southern Arizona, emphasizing vulnerability in these arid environments.

@@

Given the limited coverage provided by existing gauges, a common issue in arid and semi-arid regions globally [@krabbenhoft-2022], we utilized a machine learning approach to estimate BFI in ungauged catchments. The regional model exhibited strong predictive capability (overall R²=0.764), though systematic underprediction (negaitve pbias) was noted, particularly in monsoon-dominated and warm-dry climates. These biases suggest areas for improvement in model performance, potentially through the incorporation of additional predictors or more refined regional hydrologic representations. Although creating separate regional models (by physiographic region) was explored to enhance predictive accuracy, the state-wide model consistently outperformed these regionalized approaches, emphasizing the advantage of a comprehensive, state-wide dataset.

Elevation emerged as the most significant predictor, reinforcing its established role as a proxy for climate conditions influencing groundwater recharge. This aligns with findings from previous studies conducted in diverse landscapes, including New Zealand [@singh2018], across CONUS [@santhi2008], and globally [@beck2013]. The mixed influence observed for Soil Type A indicates complexity in infiltration dynamics, suggesting that future research could benefit from explicitly incorporating temporal variables such as snowpack duration, soil moisture, and detailed land-cover characteristics.

@@

This research highlights critical limitations in Arizona’s current streamgage network for monitoring base-flow processes, suggesting a need for targeted investments in instrumentation capable of capturing low-flow dynamics. Additionally, the demonstrated success of our modeling framework accentuates its broader applicability to similarly data-scarce dryland regions globally, providing a valuable tool for water resource management and climate adaptation strategies. By improving our understanding of groundwater-surface water interactions under changing climatic conditions, this study contributes substantially to addressing hydrological uncertainties in arid and semi-arid environments.
